---
title: "ST 431/531 Homework 1"
author: "Name"
date: "Summer 2018"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
subtitle: Due on Canvas July 27th 
---

## Problems (25 points total)

1. Identify the population, variable, and paremter of interest in the following scientific questions:
    a) (3 points) Estimate the average number of oranges on orange trees at Robertson's Farm.
    Answer: It seems similar to the OSU students in CS department where OSU Students is population and department is categorical variable
        - Population: Number of Oranges on orange trees (Oranges)
        - Variable: at Robertson's farm (categorical variable)
        - Parameter:Mean (Average)
        
    b) (3 points) Estimate the 20th percentile for weight for babies born in Oregon hospitals in 2017.
        - Population:babies born in oregon hospitals in 2017 (babies)
        - Variable:weight (Similar to length of fish--> continous variable)
        - Parameter: 20th percentile

2. Use the following code to load the **ex0116** data set containing the gross domestic product (GDP) per capita for 228 countries in 2010.

    ```{r echo=TRUE}
    library(Sleuth3)
    data(ex0116)
    ```

    a) (2 points) Create a histogram of the population probability distribution. What do you notice about the shape of the distribution?
    
    ```{r}
    library(ggplot2)
       ggplot(ex0116, aes(PerCapitaGDP))+geom_histogram(bins=30,color="blue",fill="tomato")+xlab("GDP") + ggtitle("Question 2: probability distribution")
    
    
    ```
    
        - The distribution... The sidtribution seems unbalanced and also, for example you can see that the very first bar in the histogam is the largest one and others are smaller. apart from this we can also observe that we have some data which are way away from the usual data for example small bars on the right end side.
    
    
    b) (1 point) What is the population mean GDP per capita?
    
    ```{r echo=TRUE}
    mean(ex0116[,3])
    mean(ex0116$PerCapitaGDP)
        
    ```
        
        - Population mean: 
        
    c) (2 points) Use the following code to draw a random sample of size $n=10$ from this population. What is the sample mean? What is the sample variance?
    
    ```{r echo=TRUE}
        samp1 <- sample(ex0116$PerCapitaGDP, size=10, replace=FALSE)
    mean(samp1)
    var(samp1)
    ```
    
        - Sample mean: Sample mean is the mean of the sample data
        - Sample variance: sample variance is the variance of the sample data
    
    d) (3 points) Repeat part (c) to obtain a different random sample of size $n=10$. What are the sample mean and sample variance from this sample? Why are these values different from those in part (c)?
    
    ```{r echo=TRUE}
    samp1 <- sample(ex0116$PerCapitaGDP, size=10, replace=FALSE)
    mean(samp1)
    var(samp1)
        
    ```
    
        - Sample mean: since we are selecting data randomly so the sample mean and variance will be different everytime.
        - Sample variance: since we are selecting data randomly so the sample mean and variance will be different everytime.
        - These values are different because...
    Answer: These values are different beacuse sample in previous question and in this question are different since we are taking randome 10 sample with the replacement= FALSE.
    
    

3. (4 points) Consider the following three sampling distributions for the sample average from the **ex0116** GDP data used in question 2. One distribution is obtained from samples of size $n=5$, one is obtained from samples of size $n=25$, and one is obtained from samples of size $n=100$. Which histogram (1,2,3) corresponds to which sample size? How can you tell (besides looking at the R code)? Explain.

    ```{r sampdist, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=8, fig.height=3, out.width="80%", fig.align="center"}
library(ggplot2)

d1 <- c()
d2 <- c()
d3 <- c()

for (i in 1:100000) {
  d1[i] <- mean(sample(ex0116$PerCapitaGDP, 5, replace=FALSE))
  d2[i] <- mean(sample(ex0116$PerCapitaGDP, 25, replace=FALSE))
  d3[i] <- mean(sample(ex0116$PerCapitaGDP, 100, replace=FALSE))
}

df <- data.frame(x=c(d1, d2, d3), f=rep(c(3,1,2), each=100000))

ggplot(df, aes(x=x)) + geom_histogram() + facet_grid(. ~ f) + theme_bw()
    ```

    - Histogram 1 corresponds to... n=25
    - Histogram 2 corresponds to... n=100
    - Histogram 3 corresponds to... n=5
    - This is because... figure 3 not balanced and more samples are presents whereas in the figure 1 represents the n=25 because this also seems less balanced then the figure 2 and also the samples are more than the figure 2 and hence figure 2 represents for the n=100.

4. Recall that if the population distribution has mean $\mu$ and variance $\sigma^2$, the Central Limit Theorem says that for a sample of size $n$, the sample mean has an approximately Normal distribution with mean $\mu$ and variance $\sigma^2/n$.

    a) (3 points) Suppose a population has mean $\mu=40$ and variance $\sigma^2=25$. What is the approximate distribution of the sample mean for samples of size $n=20$?
    
        - The distribution is... Normal(40, 25/20)
        
       - X ~ Normal($\mu$,$\sigma$/n) --> Normal(40, 25/20)
    
    b) (4 points) Suppose a population has mean $\mu=100$ and variance $\sigma^2=20$. For a sample of size $n=10$, what is the approximate probability that the sample mean is less than 98?
    ```{r echo=TRUE}
   pnorm(98, mean=100, sd=sqrt(2))
        
    ```
    
        - The probability is 0.0786496 
        
        
