---
title: 'ST 411/511 Lab 2: Working with the Normal Distribution in R and  t-tests'
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the packages we will use in this lab
library(ggplot2)
```

In today's lecture, you will see/have seen data on Rotten Tomatoes ratings for movies in 2015. Specifically, the list of movies is 146 films on Fandango.com that had tickets on sale in 2015, as of August 24, 2015, and had at least 30 reviews on Fandango.com.

Data are available as a .csv file at: https://github.com/fivethirtyeight/data/tree/master/fandango

Today we will see how to work with this data in R.

**Population**: 146 films that meet the eligibility criteria

**Variable of Interest**: There are different scores for each film (e.g. IMDB ratings, Fandango stars). In lecture, we looked at the Rotten Tomatoes Tomatometer score. In lab we'll look at the IMDB User Score instead. Possible values are between 0 and 10.

First, we can read in the data directly from the web. The data are in a csv file. We load the data into R using the `read.csv()` command and save it as a data frame called **movies**.

```{r}
movies <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/fandango/fandango_score_comparison.csv")
```

We can see what the column names (variables) are in the data by using the `names` command:

```{r}
names(movies)
```

First let's make a histogram of the population. The variable we are interested in is called **IMDB**. We saw how to use `ggplot()` to make a histogram last week. In this case I have added a little bit of customization -- `binwidth=0.5` means that I want each bar on the histogram to be 0.5 units wide, `color="black"` means that I want the outline of the bars to be black, `fill="blue"` means that I want the inside of the bars to be blue, and `alpha=0.5` means that I want the opacity of the bars to be 50% (this produces a lighter blue color). Try some different values for these options, perhaps different colors, bin widths, or opacities.

```{r}
ggplot(movies, aes(x=IMDB)) + geom_histogram(binwidth=0.5, color="black", alpha=0.5, fill="blue") + ggtitle("Population Histogram") + xlab("Tomatometer Score")
```

## Examining the population

We can calculate the population variance $\sigma^2$, which we need to know to perform a $z$-test. Remember that we want to take the variance of the IMDB scores, so we need to specify the name of the data frame, followed by a dollar sign, followed by the name of the column within the data frame.

```{r}
var(movies$IMDB)
```

It may also be of interest to us to examine specific values of the data. Which movie had the highest IMDB rating? The `which.max()` function will return the row number of the movie with the highest IMDB score. We then display just that row using the `[,]` notation. The number(s) before the comma are the row numbers we want to display, and the number(s) after the comma are the column numbers we want to display. Film name is the first column and IMDB is the sixth

```{r}
idmax <- which.max(movies$IMDB)
movies[idmax, c(1,6)]
```

Which movie did you get? Did you see this movie?

What about the movie with the lowest IMDB user score? We can use the `which.min()` function to find this. It has the same syntax as `which.max()`.

```{r}
idmin <- which.min(movies$IMDB)
movies[idmin, c(1,6)]
```

## Drawing a specific sample

We know that when we draw a **random sample**, we get different samples each time. This is a property we want, but it makes it difficult to discuss the results of a particular random sample. 

We can force R to draw the same random sample by using the `set.seed()` function. The argument in `set.seed()` is any number you'd like -- you can pick 12345, 352789908, your favorite actor's birthday, the fist 10 digits of pi, etc. Let's give this a try and draw a sample of size $n=20$. The vector **samp** will consist of the row numbers of the 20 randomly selected movies. The last line of code will print out just the selected rows in the data frame, showing the title and IMDB score.

(Note: because some of these titles are long, the columns may not all display together. After running this code chunk, you can use the white arrows at the top of the output to scroll between the columns. You can use the previous and next buttons at the bottom of the output to cycle between the first 10 movies and the second 10 movies.)

```{r}
set.seed(411511)
samp <- sample(1:nrow(movies), size=20)

movies[samp, c(1,6)]
```

You should get the same list of movies as I did, even though we are running the code on separate computers because we used the same seed. The first 3 movies in my sample are Welcome to Me, Unfriended, and Annie. Did you get the same ones?

Try running this code chunk again by clicking the green arrow in the upper right corner of the chunk. You should get the same sample again.

Try changing the number you gave in `set.seed()`. Now you should get a different list of movies.

## $z$-test

Let's work through the process of a $z$-test. Here are the null and alternative hypotheses we are interested in testing:

**Null hypothesis**: $H_0:\mu=6.5$. In words: the null hypothesis is that the population mean IMDB score is 6.5.

**Alternative hypothesis**: $H_A:\mu>6.5$. In words: the alternative hypothesis is that the population mean IMDB score is greater than 6.5. This is a one-sided greater alternative hypothesis.

**Test statistic**: this is our sample mean minus $\mu_0$ (the value of $\mu$ assuming the null hypothesis is true), divided by the standard deviation of the sampling distribution.

```{r}
# Make sure you are using the same seed as me if you want to check your answer.
set.seed(411511)
samp <- sample(1:nrow(movies), size=20)

Xbar <- mean(movies$IMDB[samp])
sigma2 <- var(movies$IMDB)
  
Z <- (Xbar - 6.5) / sqrt(sigma2/20)
Z
```

I got $Z=0.443$.

**p-value**: The $p$-value is the probability under the null hypothesis that we would observe the test statistic we did or something more extreme just due to chance. Because we are using a one-sided greater alternative hypothesis, "something more extreme" means values of $Z$ that are larger than 0.443. This is the area under the standard normal curve that is above 0.443. `pnorm(Z)` will give the area under the curve BELOW the value of $Z$, so if we want the area above we can just use `1-pnorm(Z)`.

```{r}
1-pnorm(Z)
```
I got a $p$-value of 0.329.

**Conclusion**: We fail to reject the null hypothesis $H_0:\mu=6.5$ at significance level $\alpha=0.05$. We do not have enough evidence to suggest that the population mean IMDB score is less than 6.5.

**Now you try:**

Repeat this procedure for the same null hypothesis, but one-sided lesser and two-sided alternative hypotheses. State what the alternative hypothesis is in each case, calculate the test statistic, find the $p$-value, and state the conclusion of the test in words.

HINT: think about whether or not the value of $Z$ will be different than above.

Check your work: the correct p-values are 0.671 for the one-sided lesser alternative hypothesis and 0.658 for the two-sided alternative hypothesis. Note that you needed to use the same `set.seed()` value as me to get the same random sample in order to get the same $p$-values.

## Want more practice?

You can try:

- Use your random sample to test a different set of hypotheses (e.g., your null could be $H_0:\mu=4.5$).
- Draw a different random sample and test the same hypotheses.
- Draw a different random sample and test different hypotheses.

## One-sample $t$-test

A one-sample $t$-test is very similar to a one-sample $z$-test, except that we do not know the population variance $\sigma^2$. We must therefore estimate $\sigma^2$ using the sample variance $s^2$. Our test statistic is therefore $t(\mu)$, and the null distribution is a $t$-distribution with $n-1$ degrees of freedom: $t_{(n-1)}$.

Suppose we are interested in the average IQ score of Corvallis residents. We draw a random sample of 8 Corvallis residents and test $H_0:\mu=100$ versus the two-sided alternative $H_A:\mu\neq 100$ at significance level $\alpha=0.1$. These are the 8 IQ scores in our sample:

```{r}
IQ <- c(90, 102, 100, 140, 68, 89, 109, 75)
var(IQ)
mean(IQ)

```

We don't know the population variance, so we will perform a $t$-test. First, let's find our $t$-statistic:

$$t(\mu_0)=\frac{\bar{X}-\mu_0}{\sqrt{s^2/n}}=\frac{96.625-100}{\sqrt{494.84/8}}=-0.429$$
$$
\frac{50}{2}=5^2
$$

```{r}
xbar <- mean(IQ)
s2 <- var(IQ)
tstat <- (xbar-100) / sqrt(s2/8)
tstat
```

This value is NEGATIVE because our observed sample mean (96.625) is smaller than $\mu_0$ (100). Values of the $t$-statistic (as with the $z$-statistic) can be between negative infinity and positive infinity. However $p$-values are probabilities, and so they can only be between 0 and 1.

Let's find the critical values for a test at significance level $\alpha=0.1$. We are doing a two-sided test, so the critical values $t_{(n-1)1-\alpha/2}$ will be the values on a $t$-distribution with $n-1$ degrees of freedom such that there is area 0.05 in each tail.

Why are we using `qt()`? 
  Because we are looking up a quantile. 

Why are we using `df=7`? 
  Because the degrees of freedom for the $t$-distribution is the sample size $n$ minus 1. In this case, the sample was of 8 people.

```{r}
qt(0.05, df=7)
qt(0.95, df=7)
```
The $t_{(7)}$-distribution is symmetric. For this distribution, the area below -1.895 is 0.05 and the area above 1.895 is 0.05, so the total area in the tails ($\alpha$) is 0.1.

How might we plot this in R? NOTE: I have included this code in case you are curious how to produce these plots. It is not required to know, and it is okay if this seems confusing. For now, look at the plot and ignore the code (unless you're curious). The plot shows the rejection region shaded in orange.

- Where are the critical values you found above?
- Where is the test statistic we calculated?
- Based on the plot, would we reject or fail to reject the null hypothesis?

```{r}
funcShaded <- function(x) {
    y <- dt(x, df=7)
    y[x > qt(0.05, df=7) & x < qt(0.95, df=7)] <- NA
    return(y)
}
ggplot(data = data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dt, n = 101, args = list(df=7)) + 
  ylab("") +
  stat_function(fun=funcShaded, geom="area", fill="orange", alpha=0.5)
```

Answers:

- The critical values (+/- 1.895) are at the inner edges of the rejection region.
- The test statistic (-0.429) is toward the middle of the plot.
- The test statistic does not fall within the rejection region (shaded orange), so we would fail to reject the null hypothesis. The average IQ score of Corvallis residents could be 100.

What is the $p$-value for this test? Note we are still using the $t$-distribution with 7 degrees of freedom here and that we were considering a two-sided test.

```{r}
2*(1-pt(abs(tstat), df=7))
```

Since $p=0.681>\alpha=0.1$, we fail to reject the null hypothesis that $\mu=100$. This is a different way of obtaining the same result as before.

I think it's good practice (and helps with conceptual understanding!) to calculate test statistics, critical values, and $p$-values manually, but there is a built-in function in R called `t.test()` you can use once you've gotten the hang of it. Let's try it out:

```{r}
t.test(IQ, mu=100)
```
We specify the vector of our data as the first argument, and then need to specify what value of $\mu_0$ we use in our null hypothesis. Take a moment to look at the output from R. The function automatically knows that our degrees of freedom for the $t$-distribution should be 7. The $t$-statistic and $p$-value we find here are also the same as before.

As we get into more complicated $t$-tests we can specify additional arguments in the `t.test()` function.

How would we test a one-sided alternative hypothesis? We can specify that using the argument `alternative="less"` or `alternative="greater"` in the `t.test()` function. Try it out! Calculate a p-value by hand for a one-sided alternative, and then try it in `t.test()` and see if you get the same value.

One-sided lesser alternative

```{r}
pt(tstat, df=7)

t.test(IQ, mu=100, alternative="less")
```

