---
title: 'ST 411/511 Lab 5: Analysis of Variance'
date: Aug 7, 2018
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the packages we will use in this lab
library(ggplot2)
library(Sleuth3)
```

## Outline

In today's lab, we will look at ANOVA and the Kruskal-Wallis Test for Case Study 1 from Chapter 5 of *The Statistical Sleuth*. We will then return to the Spock Conspiracy Trial data discussed during lecture.

## Overview of One-way Analysis of Variance (ANOVA)

Analysis of variance (ANOVA) is a method for evaluating relationships between means for *several* populations (i.e., more than 2). While it's called analysis of *variance*, it's very much a procedure used to test the equality of several *means*. In fact, if we've obtained samples from $I>2$ populations, the primary null hypothesis that we can investigate with one-way ANOVA is given by:

> $H_0: \mu_1=\mu_2=\cdots =\mu_I$

In this regard, one-way ANOVA is a generalization of the two-sample problems we've been looking at so far in class.

### Exploratory Plots: Longevity Case Study

The dataset in `case0501` in the `Sleuth3` library contains lifespan measurements (in months) for 6 groups of female mice. Each group was fed a different diet--each diet was comprised of various levels of caloric intake and/or protein content both before and after weaning. Let's take a look at the help file and the data:

```{r}
names(case0501)
?case0501
ggplot(data = case0501, aes(Diet, Lifetime)) + 
  geom_boxplot(fill = "yellow") + 
  ylab("Lifetime (months)")
```

By default in R, the levels of a factor variable, in this case the *Diet* variable are sorted alpha-numerically, with capital letters going before lower case letters. You can see the default ordering by using the `levels()` function, shown below, and verify that this is the order in which the side-by-side boxplots appear above.

```{r}
levels(case0501$Diet)
```

### Questions of Interest

In the diet and longevity case study, the researchers wanted to know which, if any, of the diets resulted in longer life spans. In Statistics we break this problem into two parts:

1. Are the population means the same or different?

2. If there is evidence that the population means are different, then we look at specific *pairwise* comparisons (i.e., comparisons between two groups at a time) to try to determine *which* population means are different and *how* they are different.

In this lab, we'll focus on the first part. What does it looks like from the side-by-side boxplots? It's pretty clear that there's a shift in the centers of some of the *samples*, but what about the centers of the *populations*?

To address this question, we use one-way analysis of variance (one-way ANOVA), which involves a quantitative comparison of variation *between the groups* and variation *within the groups*.

We will see how to perform the analysis in several different ways.

### By Hand

#### Sum of Squares WITHIN

The sum of squares within is found by summing up the squares of the observations minus their group means, $\sum_{i,j}(X_{i,j}-\bar{X}_i)^2$. As mentioned in lecture, there are faster/tidier ways of doing these calculations. I have used this version to cut down on specialized functions so it is more apparent exactly what calculations we are doing.

```{r}
# Create a vector of IDs which list which rows correspond to each diet.
ID85 <- which(case0501$Diet == "N/N85")
ID40 <- which(case0501$Diet == "N/R40")
ID50 <- which(case0501$Diet == "N/R50")
IDNP <- which(case0501$Diet == "NP")
IDR <- which(case0501$Diet == "R/R50")
IDlo <- which(case0501$Diet == "lopro")

# Create a vector of lifetime means by group
groupmeans <- c(mean(case0501$Lifetime[ID85]), mean(case0501$Lifetime[ID40]), mean(case0501$Lifetime[ID50]), mean(case0501$Lifetime[IDNP]), mean(case0501$Lifetime[IDR]), mean(case0501$Lifetime[IDlo]))

SSW <- sum((case0501$Lifetime[ID85]-groupmeans[1])^2) +
  sum((case0501$Lifetime[ID40]-groupmeans[2])^2) +
  sum((case0501$Lifetime[ID50]-groupmeans[3])^2) +
  sum((case0501$Lifetime[IDNP]-groupmeans[4])^2) +
  sum((case0501$Lifetime[IDR]-groupmeans[5])^2) +
  sum((case0501$Lifetime[IDlo]-groupmeans[6])^2)
SSW

```

This is also the sum of squared residuals under the *full/separate means* model.

#### Sum of squares TOTAL

The sum of squares TOTAL is found by summing up the squares of the observations minus the grand mean, $\sum_{i,j}(X_{i,j}-\bar{X})^2$.

```{r}
gm<-(sum(case0501$Lifetime[ID85])+ sum(case0501$Lifetime[ID40])+ sum(case0501$Lifetime[ID50])+ sum(case0501$Lifetime[IDNP])+ sum(case0501$Lifetime[IDR])+ sum(case0501$Lifetime[IDlo]))
mean(case0501$Lifetime[ID85])
sum(case0501$Lifetime[ID85])
gm
grandmean <- mean(case0501$Lifetime)
SST <- sum((case0501$Lifetime - grandmean)^2)
SST
```

This is also the sum of squared residuals under the *reduced/equal means* model.

#### Sum of squares BETWEEN

We can find the sum of squares BETWEEN by taking the sum of squares TOTAL minus the sum of squares WITHIN, SSB=SST-SSW.

```{r}
SSB <- SST - SSW
SSB
```

Note that this is also the *Extra sum of squares*.

#### Degrees of freedom

We know how many total observations there are and how many groups there are, so we can find the degrees of freedom.

```{r}
n <- nrow(case0501)
n
I <- 6
dfW <- n - I
dfT <- n - 1
dfB <- I - 1
```

Note that the degrees of freedom BETWEEN is also the *Extra degrees of freedom*.

#### Mean squares

The mean square is the sum of squares divided by the corresponding degrees of freedom.

```{r}
MSW <- SSW / dfW
MSB <- SSB / dfB
```

#### $F$-statistic

The $F$-statistic is the mean square BETWEEN divided by the mean square WITHIN:

```{r}
F <- MSB / MSW
F
```

We also saw that we could think of it as the (extra sum of squares/extra degrees of freedom), divided by the pooled variance, where we know the pooled variance is equal to the mean square WITHIN. You can calculate this to show it's the same.

#### $p$-value

We want to look up the area in the F distribution with dfW and dfB degrees of freedom that is above our observed $F$-statistic.

```{r}
1-pf(F, df1=dfB, df2=dfW)
```

This $p$-value is very, very small -- so close to zero that 0 gets printed out due to rounding. We would reject the null hypothesis that the population mean lifetimes are all the same. It appears that at least one of the population mean lifetimes is different from the others.

### Using the `anova` function wrapped around the `lm` function:

We use the same *formula* syntax we have looked at for other functions, where we specify the names of the columns in our data that correspond to the response variable ~ the grouping variable.

```{r}
anova(lm(Lifetime ~ Diet, data=case0501))
```

Do these values agree with what you calculated by hand? Check a few values. Note that *Diet* is our between group, and *Residuals* is our within group.

### Using the `aov` function and then calling `summary` to print out the ANOVA table:

```{r}
summary(aov(Lifetime ~ Diet, data=case0501))
```

Are these values the same? What about if you just use `aov` without using `summary`?

## Kruskal-Wallis Test

The *Kruskal-Wallis Test* is a multiple group analog of Wilcoxon Rank-Sum Test. We can perform it in R using the `kruskal.test` function. The syntax is again the same formula syntax we are used to.

```{r}
kruskal.test(Lifetime ~ Diet, data=case0501)
```

Note that this $p$-value is also very small!

## More on the Spock Conspiracy Trial

We've looked through this data in lecture, but we can dive a little deeper here. 

Let's quickly look at the data and perform the analysis of variance again:

```{r}
ggplot(case0502, aes(x=Judge, y=Percent)) + geom_boxplot()

anova(lm(Percent ~ Judge, data=case0502))
```

We remember that our ANOVA provided strong evidence that not all of the judges had the same percent women on their venires, with $p$-value = 0.00006. We only have tested whether or not all of the population means are the same -- we haven't tested which one(s) are different. From examining the boxplot, it seems that the Spock's judge was different, but we have not stated that statistically.

There are several approaches we could try based on what we've learned so far. Here are two questions we could look at:

1. Does Spock's judge mean percentage differ from all the others?
2. Are the other judges mean percentages the same?

#### Does Spock's judge mean percentage differ from all the others?

To address this question, let's fit a one-way ANOVA model to see what evidence there is that the Spock trial judges venires are different from all of the other judges' venires combined (this is equivalent to a two-sample t-test).

Note that `(Judge == "Spock's")` creates a TRUE/FALSE outcome for whether or not the judge is Spock's, so now we are modeling Percent based on just two groups, not seven.

```{r}
anova(lm(Percent ~ (Judge == "Spock's"), data = case0502))
```

This gives convincing evidence that the mean percent of women on the Spock judge's venires is different from the means of all the other judges' venires combined.  To see the equivalence with the two-sample $t$-test consider the following:

```{r}
t.test(Percent ~ (Judge == "Spock's"), data = case0502, var.equal = TRUE)
```

Notice that the $p$-value we get from the two-sample $t$-test with equal variance is exactly the same as what we get from the one-way ANOVA comparing the two groups. Also, notice that the square of the $t$-statistic is equal to the F-statistic.

#### Are the other judges mean percentages the same?

Let's fit a one-way ANOVA model *to Judges A, B, ..., F venires only* to see what evidence there is for all six of those underlying means being the same. We will look at whether there's evidence that the six judges other than the Spock trial judge differ in terms of the mean percent of women on the venires. We can use the `subset` function to look at only observations of the data where the judge is not Spock's.

```{r}
anova(lm(Percent ~ Judge, data = subset(case0502, Judge != "Spock's")))
```

Here, there's no evidence of a difference in the mean percent of women on the venires of these six judges ($p = 0.32$). 

Putting this altogether, it's not just that the Spock trial judge had the lowest mean percent of women, it's also that none of the other judges had substantially different mean percents of women. Combined, this seems like compelling evidence of a bias against women on the part of the Spock trial judge.

